Sexual selection, environmental variation, range size and speciation in passerine

### 2018-10-10 Manuscript.Rmd
Edits and added data objects needed for recompilation + figures

---
title: Sexual selection, environmental variation, range size and speciation in passerine
  birds
author:
- Justin G. Cally*\textsuperscript{§}, Devi Stuart-Fox\textsuperscript{§}, Luke Holman\textsuperscript{§} and Iliana Medina Guzmán\textsuperscript{§}
- ~
- '*\textit{justin.g.cally@gmail.com}'
bibliography: References/SS_Extinction.bib
output:
  pdf_document: default
  html_document: default
  word_document:
    pandoc_args: --smart
header-includes: \usepackage{microtype} \usepackage[all]{hypcap} \usepackage{amsmath}
  \usepackage{booktabs} \usepackage{caption} \usepackage[labelfont=bf]{caption} \usepackage{titling}
  \pretitle{\begin{flushleft}\LARGE} \posttitle{\par\end{flushleft}\vskip 0.5em} \preauthor{\begin{flushleft}\large
  \lineskip 0.5em} \postauthor{\par\end{flushleft}} \predate{\begin{flushleft}\large}
  \postdate{\par\end{flushleft}} \usepackage{fancyhdr} \pagestyle{fancy} \fancyhead[LO,LE]{\textsl{\leftmark}}
  \rhead[]{Sexual Selection and Speciation} \usepackage[noabbrev,capitalise]{cleveref}
  \usepackage{titlefoot} \usepackage{amssymb} \usepackage{rotating} \usepackage{lineno}
  \linenumbers \usepackage{setspace}\onehalfspacing \usepackage[font={small}]{caption}
  \usepackage{epigraph} \setlength\epigraphwidth{5.25in} \setlength\epigraphrule{0pt}
  \setlength\beforeepigraphskip{1\baselineskip} \setlength\afterepigraphskip{2\baselineskip}
  \renewcommand*{\textflush}{flushright} \renewcommand*{\epigraphsize}{\footnotesize\itshape}
  \usepackage{hyperref}\hypersetup{colorlinks=true, linkcolor=blue, filecolor=blue, urlcolor=blue}
  \raggedbottom
csl: References/AmNat.csl
abstract: 'Sexual selection may influence patterns of diversification; yet controversy remains on whether sexual selection leads to a net increase or decrease in speciation rate, or even elevates extinction. Sexual selection is predicted to increase speciation rate as it can facilitate adaptation to new environments, and work alongside divergent natural selection to swiftly promote reproductive isolation. However, sexual selection frequently favours investment in costly or harmful traits that may reduce adaptation and affect diversification. Here, we investigate the association of sexual selection with speciation and extinction in passerine birds. We use two measures of sexual selection: sexual dichromatism (n = 5,812 species) and a multivariate measure of male-biased sexual selection (sexual dimorphism, polygyny and paternal care; n = 2,465), three distinct measures of speciation ($\lambda_{DR}$, $\lambda_{ND}$, $\lambda_{BAMM}$) and one measure of extinction ($\mu_{BAMM}$). Importantly, we test whether the effects of sexual selection on speciation are dependent upon ecological pressures such as temperature variability across time and space, primary productivity and range size. We find that male-biased sexual selection but not sexual dichromatism predicts speciation. Additionally, we find that birds with smaller ranges have higher speciation rates; although the direction of causality is unclear.'
---
\maketitle\unmarkedfntext{\textsuperscript{§}School of BioSciences, The University of Melbourne, Victoria, Australia.}

@@

Selection pressures are the main engine of speciation. As populations diverge they accumulate genetic and phenotypic changes that ultimately erect barriers against gene flow [@Presgraves_2010; *but see* @Shaw_2011]. In this view, populations will speciate and adapt more readily to new and different environments if selection is strong and/or persistant. Hence, the rate at which speciation occurs is likely dependent on both the populations response to selection (adaptation) and the potential for divergent phenotypic traits to evolve [*e.g.* morphology or colour @Hugall_2012; @Rabosky_morph_2013]. Sexual selection is an evolutionary force likely to affect fitness components of a population, such as reproductive success or viability, as well as the evolution of ornaments involved in mating displays or weapons used in male-male competition (**add couple of refs**). Sexual selection is therefore hypothesised to be a key driver of diversification through shifting the rates of adaptation, speciation and extinction.

@@

Alongside changes to fitness, sexual selection has been proposed to facilitate speciation through the evolution of divergent signals/ornaments associated with improved mating success [@Lande_1981; @Lande_1982]. For instance, in antbirds (Thamnophilidae), genera with complex songs and striking dichromatism, are more species rich [@Seddon_2008]. Under this model of speciation, divergent sexual signals in one sex (usually males) co-evolve with divergent preferences for those signals in the opposite sex, leading to behavioural reproductive isolation [@Safran_2013]. Concurrently, sexual selection can increase the extent to which populations diverge by enlarging the phenotypic space diverging species can evolve into. These predictions emerge as sexual selection may increase genetic diversity as a result of trade-offs between naturally- and sexually-selected traits or trade-offs under intra-locus sexual conflict between male and female trait expressions [@Lorch_2003; @Bonduriansky_2011; @Radwan_2016]. However, the genetic architecture (*per se* heritability) of fitness and sexually selected traits may vary due to different sex determination systems across taxa. In male homogametic species (ZZ/ZW or ZZ/ZO) --- such as birds --- males are predicted to have increased heritability of male-specific genotypes on the Z chromosome associated with mating success: such as elaborate ornamentation or plumage colouration [@Hastings_1994; @Reeve_2003]. We might therefore expect bird populations experiencing stronger sexual selection to have increased adaptation and speciation rates. **I think the link between the increased heritability of of the ornamentation and adaptation and speciation should be explained more clearly**

Importantly, the effects of sexual selection on adaptation and speciation may depend on environmental conditions. Both theoretical and empirical studies suggest that under stressful/changing environments, sexual selection may have greater fitness benefits than under benign environments. Stressful environments strengthen the positive correlation between male mating success and female fitness, reducing the burden of negative pleiotropic effects [@Berger_2014; @Connallon_Hall_2016; @Long_2012; @Martinossi-Allibert_Savković_2017]. Conversely, benign or stable environments ensure consistent selection, preferentially eroding genetic variation at sexually concordant loci (*i.e.* loci where the optimum genotype is equivelant for both sexes). In this situation, genetic variation remains disproportionately at sexually antagonistic loci; causing displacement of male and female traits from their optimum. It follows from these predictions that populations with stronger sexual selection would have increased fitness and adaptive benefits, allowing them to adapt to novel environments [*reviewed in* @Candolin_2008]. In certain circumstances, sexual selection will reinforce locally adaptated phenotypes through mate choice on phenotypes that effectively advertise quality in a specific ecological context [@Boughman_2002; @Maan_2011]. Additionally, if mate choice operates on both sexes --- or there is a correlation in trait values between a pair --- assortative mating is likely to speed-up the evolution of divergent mating signals [@Lande_1981]. In several bird taxa, beak morphology is a trait that arose under natural selection and is now a trait under sexual selection [a.k.a. "magic traits"; @Servedio_2011]. In Darwin's finches (*Geospiza fortis*) divergent beak morphology is an adaptation to local food availability that has been maintained through assortative mating [@Huber_2007]. Alternatively, sexual selection may also interact with natural selection in opposing directions when environmental conditions modify the sexually selected signal, such as turbid waters that reduce the signal efficacy in sticklebacks [@Wong_2007; @Candolin_2007]. In this conflicting view, changing environments may limit the divergence of sexually selected traits and consequently reduce speciation rate. *** would be good to have in this paragraph a bit of how the unstability of the environment can be viewed as a proxy for harsher conditions, why we think that unstable environments impose larger selective pressures**

**we need a sentence earlier showing that we are doing this in birds, may be at the beginning of this paragraph, i added something**
Birds are notorious for the wide arrange of ornaments that have presumably evolved in response to sexual selection, such as plumage colour, size dimorphism, complex courtship rituals and male song (**add ref**). The association between sexual selection and speciation has been investigated previously in birds using macroevolutionary approaches. A meta-analysis of 64 effect sizes from 20 studies found a small, but significant positive association between sexual selection and speciation across a wide range of taxa [@Kraaijeveld_2011]. Within birds, they found a positive association but much lower than that for insects and fish. However, there was large variation in effect sizes due to variation in the measures of sexual selection used and diversification proxies (*e.g.* species/subspecies richness *vs* rate estimates). Contrary to this, it was recently shown by Huang and Rabosky and Cooney et al. **CITE: Sexual selection, speciation, and constraints on geographical range overlap in birds** that there is no link between sexual dichromatism and speciation in a sample of ~1000 species and ~1300 pairs of species, respectively. Given these contrasting results, and given that species richness and speciation rate are not necessarily correlated [*e.g.* @Rabosky_fish_2018] and the recent refinement of methods to obtain tip-rate diversification estimates [@Jetz_2012; @Rabosky_2014; @Rabosky_diversification_2018], the association between sexual selection and speciation rate is worth revisiting. Firstly, the strength of sexual selection in macroevolutionary studies is often estimated using measures of sexual dimorphism or dichromatism; but these are often partial measures of sexual selection and may not account for the true variation in sexual selection across taxa. Secondly, until recently, studies on the association between sexual selection and speciation have largely ignored the contribution and interaction with environmental variables to explain variation in speciation rates. Thirdly, many studies use incomplete taxon sampling, with family-level analysis restricted to well sampled avian clades [@Seddon_2008; @Huang_2014].Fourthly, phylogenetic uncertainty in the avian tree [@Jetz_2012; *but see* @Rubolini_2015] and uncertainty in tip-rate measures of speciation such as the diversification rate statistic ($\lambda_{DR}$) or node density ($\lambda_{ND}$) [@Rabosky_diversification_2018] as well as speciation and extinction tip-rates from models using BAMM (Bayesian Analysis of Macroevolutionary Mixtures) [*see*, @Beaulieu_2015; @Rabosky_Extinction_2016; @Moore_2016; @Rabosky_BAMM_flawed_2017] continue to plague efforts to identify drivers of speciation and extinction.

Here, we perform a macroevolutionary analysis of the relationship between sexual selection and speciation or extinction in passerine birds (Order: Passeriformes), arguably the largest group for which there exists comprehensive data. Recently, @Huang_2014 found no association between sexual dichromatism and speciation using tip-rate estimates from a genetic-only phylogeny [@Jetz_2012] and sexual dichromatism data from a limited spectrophotometry dataset [*n* = 918; @Armenta_2008] ** mention here the Cooney et al. paper ,or move this bit up where I added it earlier**. We build on these previous studies by using published datasets of sexual dichromatism measured from images [*n* = 5,812; @Dale_2015] alongside smaller datasets of sexual dichromatism measured from spectrophotometry [*n* = 581; @Armenta_2008] and a composite measure of male-biased sexual selection [*n* = 2,465; @Dale_2015]. We assess the relationship between sexual selection, environmental variability and their interaction with three tip-rate measures of speciation and one measure of extinction ($\lambda_{DR}$, $\lambda_{ND}$, $\lambda_{BAMM}$ and $\mu_{BAMM}$). Using expert range maps and bioclimatic variables we are able to investigate whether the effect of sexual selection on speciation is dependent on range size, seasonal variation in temperature, spatial temperature variation (across a species’ range), long-term variation in temperature between now and the last inter-glacial (LIG) as well net primary productivity (NPP).

@@

We found no evidence that sexual dichromatism alters the rate of speciation in passerine birds independently or via interactions with environmental variability. This result is consistent across different measures of speciation ($\lambda_{DR}$, $\lambda_{ND}$ and $\lambda_{BAMM}$) and two measures of dichromatism (spectral and RGB), and is consistent with previous studies performed at a smaller scale (Huang & Rabosky, Cooney et al.). However, we did find that when using a measure of male-biased sexual selection --- instead of sexual dichromatism --- there was a positive association with speciation, though the strength of this relationship varied across different rate estimates. Our findings suggest that the components of the composite measure of male-biased sexual selection (sexual size dimorphism, social polygyny and [lack of] paternal care) are better predictors of speciation than sexual dichromatism. No bioclimatic measures of environmental variability (temperature seasonality, long-term temperature variation and spatial temperature variation) predicted speciation, but range size was negatively associated with two measures of speciation ($\lambda_{DR}$, $\lambda_{ND}$). This suggests that species with smaller range sizes have increased speciation rate, however, small range size may be a cause or effect of elevated speciation rate and potentially even an artifact of taxonomic classification. Interestingly we found a discrepancy in the model results obtained from tip-rate statistics ($\lambda_{DR}$, $\lambda_{ND}$) and those using tip-rate estimates from a model-based approach ($\lambda_{BAMM}$); whereby BAMM produced precisely null results for most associations between predictors and speciation as well as extinction.

The association between sexual selection and sexual dichromatism in birds is a Darwinian trope [@Darwin_1871] that allows it to be commonly used as a proxy for the strength of sexual selection in comparative studies [*e.g.*, @Barraclough_1995; @Seddon_2013; @Morrow_Speciation_2003; @Owens_1999; @Huang_2014]. Despite the association between sexual selection and sexual dichromatism in passerine birds [@Dale_2015] we find that sexual dichromatism does not predict speciation rate, whereas male-biased sexual selection does. Given our results, we suggest that sexual dichromatism may not be a robust proxy for sexual selection, and that variance in the association between dichromatism and the strength of sexual selection may obscure any relationship with speciation rate, despite the high power of our comparative study. There are several reasons why the use of sexual dichromatism as a proxy for sexual selection is problematic. Firstly, ecological pressures could drive sexual dichromatism in the absence of mate choice if sex-specific niches provide opportunity for the evolution of sex-limited traits [@Wallace_1889; @Kottler_1980; @Slatkin_1984; @Shine_1989]. **In damselflies, differences between the sexes in their exposure to UV radiation drives dichromatism between the sexes [@Cooper_2010; @Punzalan_2010] I would erase this example and find stuff on birds** . Natural selection on females maybe important in explaining patterns of sexual dichromatism in birds ** cite: Hofmann CM, Cronin TW, Omland KE. Evolution of sexual dichromatism. 1. Convergent losses of elaborate female coloration in New World Orioles (Icterus spp.). The Auk. 2008;125:778–89.Price JJ, Eaton MD. Reconstructing the evolution of sexual dichromatism: current color diversity does not reflect past rates of male and female change. Evolution. 2014;68(7):2026–37.** In fairy-wrens (Malurus cyaneus), female colouration has evolved in response to environmental pressures, leading to increased sexual dichromatism in regions where females are under stronger selection from predation (**cite: Medina et al. https://doi.org/10.1186/s12862-016-0861-3**).  Ecological impacts on sexual dichromatism are supported by our path analysis, which reveals that sexual dichromatism is positively affected by temperature seasonality (a measure of environmental variation) albeit relatively weakly. Secondly, colour is but one trait and sexual selection may drive the evolution of sex differences in a wide variety of traits used in mate choice or intrasexual competition **cite: https://besjournals.onlinelibrary.wiley.com/doi/abs/10.1111/1365-2656.12824**. And thirdly, not all plumage colouration honestly reflects mate quality. A recent meta-analysis found that converted carotenoids (but not those sourced from diet) predict mate quality through improvements in parasite resistance and reproductive success [@Weaver_2018]. Given the importance of honest signaling in sexual selection and the variability in colour production across birds, sexual dichromatism in many species may be an unreliable measure of mate quality for female birds and sexual selection for researchers.

In-line with predictions and a previous meta-analysis [@Kraaijeveld_2011] we found that male-biased sexual selection increases speciation rate --- at least for $\lambda_{DR}$ and $\lambda_{ND}$ --- but this association was not dependent on variable or stressful environments. When evolving under stressful conditions, sexually antagonistic selection is hypothesised to be weakened relative to selection that is concordant between the sexes [@Connallon_SA_2015; @Connallon_Hall_2016]. While these predictions are supported in several lab (experimental evolution) studies [@Long_2012; @Berger_2014; @Punzalan_2013], other studies have not found the association [@Delcourt_2009; @Martinossi-Allibert_Savković_2017; @Holman_Jacomb_2017]. Given that we did not detect an effect of environmental variability (by itself or as an interaction with sexual selection) we are left with several possible explanations. ** would be good to at least mention how/why we expected one of the environmenta variables to have an effect on speciation, up to here there are not much details in the causal relationship we were expecting with environment** Firstly, the effects of sexual selection on adaptation and thus speciation may depend on the type of environmental variability the species is evolving under. Specifically, sexual antagonism may be more easily purged in environments where cyclic pressures (*e.g.* seasonality) are perpetuated; whereas when environmental variability is directional (*e.g.* long-term climatic change) sexual antagonism can remain indefinitely as a fitness burden [@Connallon_Hall_2016]. Secondly, the environmental predictors used here may not account for the key ecological forces/natural selection pressures that interact with sexual selection to drive speciation. Specifically, access to dietary resources, and the impacts of predation or parasitism are unaccounted for here: arguably these are key processes affecting sexual selection and speciation [*reviewed in* @Maan_2011]. Thirdly, it is also possible that there is no effect of environmental variability on speciation rates; at least not in the species investigated here.

@@

The most consistent finding uncovered in this comparative analysis is that smaller range sizes are associated with increased speciation rates; however reduced range size can be a cause or consequence of speciation. ** compare the effects?? looks like association with sexual selection is stronger than that with range size, which is interesting**  Intuitively, large range size should promote speciation by creating greater opportunities for geographic barriers to form [@Rosenzweig_1995]. However, **why however? seems in line with previous sentence** birds with smaller range sizes are also predicted to have lower dispersal ability. This ensues that individuals might not be able to overcome geographical barriers and maintain gene flow; which would promote reproductive isolation and speciation [@Birand_2012; @Claramunt_2012]. Several studies have used morphological measures of wing shape (Hand-wing index/ Kipp's distance) --- a proxy for dispersal ability --- as predictors of speciation and coexistence [*e.g.*, @Kennedy_2016; @Pigot_2016]. But currently, published datasets of these measurements are not available for many passerines. Alternatively, reduced range size may be a consequence of speciation. Under this view, high speciation rate may lead to smaller range sizes as niche filling by recently diverged species will suppress the expansion of newly speciated relatives. Within Himalayan songbirds, range expansions are prevented by the competition for resources by related species [@Price_2014]; however, this limitation on range expansion is then expected to prevent future speciation events [@Rosenzweig_1995; @Weir_2011]. Unlike Himalayan songbirds, species undergoing adaptive radiation in new niches are unlikely to be limited by competition for resources from existing taxa. Across islands we expect to see a correlation between speciation and small range size, not because of any inherent properties of the range size, but because they reflect newly formed islands with empty niches. Thus, our findings may be heavily dependent on species whose range size is restricted to a single island within an archipelago (*e.g.* *Geospiza* sp.; a.k.a. Darwin's finches). One further explanation for the negative association between range size and sexual dichromatism/sexual selection is the potential bias of taxonomic classification, whereby over-splitting of species in clades with large ranges leads to increased recent phylogenetic branching as well as smaller ranges.

>>>>

---
title: Sexual selection, environmental variation, range size and speciation in passerine
  birds
author:
- Justin G. Cally*\textsuperscript{§}, Devi Stuart-Fox\textsuperscript{§}, Luke Holman\textsuperscript{§} and Iliana Medina Guzmán\textsuperscript{§}
- ~
- '*\textit{justin.g.cally@gmail.com}'
bibliography: References/SS_Extinction.bib
output:
  word_document:
    pandoc_args: --smart
  html_document: default
  pdf_document: default
header-includes: \usepackage{microtype} \usepackage[all]{hypcap} \usepackage{amsmath}
  \usepackage{booktabs} \usepackage{caption} \usepackage[labelfont=bf]{caption} \usepackage{titling}
  \pretitle{\begin{flushleft}\LARGE} \posttitle{\par\end{flushleft}\vskip 0.5em} \preauthor{\begin{flushleft}\large
  \lineskip 0.5em} \postauthor{\par\end{flushleft}} \predate{\begin{flushleft}\large}
  \postdate{\par\end{flushleft}} \usepackage{fancyhdr} \pagestyle{fancy} \fancyhead[LO,LE]{\textsl{\leftmark}}
  \rhead[]{Sexual Selection and Speciation} \usepackage[noabbrev,capitalise]{cleveref}
  \usepackage{titlefoot} \usepackage{amssymb} \usepackage{rotating} \usepackage{lineno}
  \linenumbers \usepackage{setspace}\onehalfspacing \usepackage[font={small}]{caption}
  \usepackage{epigraph} \setlength\epigraphwidth{5.25in} \setlength\epigraphrule{0pt}
  \setlength\beforeepigraphskip{1\baselineskip} \setlength\afterepigraphskip{2\baselineskip}
  \renewcommand*{\textflush}{flushright} \renewcommand*{\epigraphsize}{\footnotesize\itshape}
  \usepackage{hyperref}\hypersetup{colorlinks=true, linkcolor=blue, filecolor=blue, urlcolor=blue}
  \raggedbottom
csl: References/AmNat.csl
abstract: 'Sexual selection may influence patterns of diversification; yet controversy remains on whether sexual selection leads to a net increase or decrease in speciation rate, or even elevates extinction. Sexual selection is predicted to increase speciation rate as it can facilitate adaptation to new environments, and work alongside divergent natural selection to swiftly promote reproductive isolation. However, sexual selection frequently favours investment in costly or harmful traits that may reduce adaptation and affect diversification. Here, we investigate the association of sexual selection with speciation and extinction in passerine birds. We use two measures of sexual selection: sexual dichromatism (n = 5,812 species) and a multivariate measure of male-biased sexual selection (sexual dimorphism, polygyny and paternal care; n = 2,465), three distinct measures of speciation ($\lambda_{DR}$, $\lambda_{ND}$, $\lambda_{BAMM}$) and one measure of extinction ($\mu_{BAMM}$). Importantly, we test whether the effects of sexual selection on speciation are dependent upon ecological pressures such as temperature variability across time and space, primary productivity and range size. We find that male-biased sexual selection but not sexual dichromatism predicts speciation. Additionally, we find that birds with smaller ranges have higher speciation rates; although the direction of causality is unclear.'
---
\maketitle\unmarkedfntext{\textsuperscript{§}School of BioSciences, The University of Melbourne, Victoria, Australia.}

@@

Selection pressures are the main engine of speciation. As populations diverge they accumulate genetic and phenotypic changes that ultimately erect barriers against gene flow [@Presgraves_2010; *but see* @Shaw_2011]. In this view, populations will speciate and adapt more readily to new and different environments if selection is strong and/or persistant. Hence, the rate at which speciation occurs is likely dependent on both the populations response to selection (adaptation) and the potential for divergent phenotypic traits to evolve [*e.g.* morphology or colour @Hugall_2012; @Rabosky_morph_2013]. Sexual selection is an evolutionary force likely to affect fitness components of a population [@Cally_in_review], such as reproductive success or viability, as well as the evolution of ornaments involved in mating displays or weapons used in male-male competition [@McCullough_2016]. Sexual selection is therefore hypothesised to be a key driver of diversification through shifting the rates of adaptation, speciation and extinction.

@@

Alongside changes to fitness, sexual selection has been proposed to facilitate speciation through the evolution of divergent signals/ornaments associated with improved mating success [@Lande_1981; @Lande_1982]. For instance, in antbirds (Thamnophilidae), genera with complex songs and striking dichromatism, are more species rich [@Seddon_2008]. Under this model of speciation, divergent sexual signals in one sex (usually males) co-evolve with divergent preferences for those signals in the opposite sex, leading to behavioural reproductive isolation [@Safran_2013]. Concurrently, sexual selection can increase the extent to which populations diverge by enlarging the phenotypic space diverging species can evolve into. These predictions emerge as sexual selection may increase genetic diversity as a result of trade-offs between naturally- and sexually-selected traits or trade-offs under intra-locus sexual conflict between male and female trait expressions [@Lorch_2003; @Bonduriansky_2011; @Radwan_2016]. However, the genetic architecture (*per se* heritability) of fitness and sexually selected traits may vary due to different sex determination systems across taxa. In male homogametic species (ZZ/ZW or ZZ/ZO) --- such as birds --- males are predicted to have increased heritability of male-specific genotypes on the Z chromosome associated with mating success: such as elaborate ornamentation or plumage colouration [@Hastings_1994; @Reeve_2003]. If fitness and associated mating traits are more heritable in birds we might expect bird populations experiencing stronger sexual selection to have increased adaptation and speciation rates compared to male heterogametic taxa.

Importantly, the effects of sexual selection on adaptation and speciation may depend on environmental conditions. Both theoretical and empirical studies suggest that under stressful/changing environments, sexual selection may have greater fitness benefits than under benign environments. Stressful environments strengthen the positive correlation between male mating success and female fitness, reducing the burden of negative pleiotropic effects [@Berger_2014; @Connallon_Hall_2016; @Long_2012; @Martinossi-Allibert_Savković_2017]. Conversely, benign or stable environments ensure consistent selection, preferentially eroding genetic variation at sexually concordant loci (*i.e.* loci where the optimum genotype is equivelant for both sexes). In this situation, genetic variation remains disproportionately at sexually antagonistic loci; causing displacement of male and female traits from their optimum. It follows from these predictions that populations with stronger sexual selection would have increased fitness and adaptive benefits, allowing them to adapt to novel environments [*reviewed in* @Candolin_2008]. In certain circumstances, sexual selection will reinforce locally adaptated phenotypes through mate choice on phenotypes that effectively advertise quality in a specific ecological context [@Boughman_2002; @Maan_2011]. Additionally, if mate choice operates on both sexes --- or there is a correlation in trait values between a pair --- assortative mating is likely to speed-up the evolution of divergent mating signals [@Lande_1981]. In several bird taxa, beak morphology is a trait that arose under natural selection and is now a trait under sexual selection [a.k.a. "magic traits"; @Servedio_2011]. In Darwin's finches (*Geospiza fortis*) divergent beak morphology is an adaptation to local food availability that has been maintained through assortative mating [@Huber_2007]. Therefore, strong sexual selection and strong natural selection --- such as environmental change over space and time --- may work synergestically to increase speciation rates.

<!-- Alternatively, sexual selection may interact with natural selection in opposing directions when environmental conditions modify the sexually selected signal, such as turbid waters that reduce the signal efficacy in sticklebacks [@Wong_2007; @Candolin_2007]. In this conflicting view, changing environments may limit the divergence of sexually selected traits and consequently reduce speciation rate.   -->

Birds are notorious for the variety of ornaments that have evolved in response to sexual selection, such as plumage colour, size dimorphism, complex courtship rituals and male song [@Gontard_1999]. The association between sexual selection and speciation has been investigated previously in birds using macroevolutionary approaches. A meta-analysis of 64 effect sizes from 20 studies found a small, but significant positive association between sexual selection and speciation across a wide range of taxa [@Kraaijeveld_2011]. Within birds, they found a positive association but much lower than that for insects and fish. However, there was large variation in effect sizes due to variation in the measures of sexual selection used and diversification proxies (*e.g.* species/subspecies richness *vs* rate estimates). Contrary to this, @Huang_2014 found no association between sexual dichromatism and speciation using tip-rate estimates from a genetic-only phylogeny [@Jetz_2012] and sexual dichromatism data from a limited spectrophotometry dataset [*n* = 918; @Armenta_2008]; similarly @Cooney_2017 found no effect of sexual dichromatism on diversification rates when using 1,306 pairs of species and human estimates of dichromatism. Given methodological advancements in tip-rate estimates [@Rabosky_diversification_2018], the association between sexual selection and speciation is worth revisiting in order to overcome previous limitations. Firstly, the strength of sexual selection in macroevolutionary studies is often estimated using measures of sexual dimorphism or dichromatism; but these are often partial measures of sexual selection and may not account for the true variation in sexual selection across taxa. Secondly, until recently, studies on the association between sexual selection and speciation have largely ignored the contribution and interaction with environmental variables to explain variation in speciation rates. Thirdly, many studies use incomplete taxon sampling, with family-level analysis restricted to well sampled avian clades [@Seddon_2008; @Huang_2014].Fourthly, phylogenetic uncertainty in the avian tree [@Jetz_2012; *but see* @Rubolini_2015] and uncertainty in tip-rate measures of speciation such as the diversification rate statistic ($\lambda_{DR}$) or node density ($\lambda_{ND}$) [@Rabosky_diversification_2018] as well as speciation and extinction tip-rates from models using BAMM (Bayesian Analysis of Macroevolutionary Mixtures) [*see*, @Beaulieu_2015; @Rabosky_Extinction_2016; @Moore_2016; @Rabosky_BAMM_flawed_2017] continue to plague efforts to identify drivers of speciation and extinction.

Here, we perform a macroevolutionary analysis of the relationship between sexual selection and speciation or extinction in passerine birds (Order: Passeriformes), arguably the largest group for which there exists comprehensive data. We build on these previous studies by using published datasets of sexual dichromatism measured from images [*n* = 5,812; @Dale_2015] alongside smaller datasets of sexual dichromatism measured from spectrophotometry [*n* = 581; @Armenta_2008] and a composite measure of male-biased sexual selection [*n* = 2,465; @Dale_2015]. We assess the relationship between sexual selection, environmental variability and their interaction with three tip-rate measures of speciation and one measure of extinction ($\lambda_{DR}$, $\lambda_{ND}$, $\lambda_{BAMM}$ and $\mu_{BAMM}$). Using expert range maps and bioclimatic variables we are able to investigate whether the effect of sexual selection on speciation is dependent on range size, seasonal variation in temperature, spatial temperature variation (across a species’ range), long-term variation in temperature between now and the last inter-glacial (LIG) as well net primary productivity (NPP).

@@

We found no evidence that sexual dichromatism alters the rate of speciation in passerine birds independently or via interactions with environmental variability. This result is consistent across different measures of speciation ($\lambda_{DR}$, $\lambda_{ND}$ and $\lambda_{BAMM}$) and two measures of dichromatism (spectral and RGB). The results are also consistent with previous studies performed at a smaller scale [@Huang_2014; @Cooney_2017]. However, we found that when using a measure of male-biased sexual selection --- instead of sexual dichromatism --- there was a positive association with speciation, though the strength of this relationship varied across different rate estimates. Our findings suggest that the components of the composite measure of male-biased sexual selection (sexual size dimorphism, social polygyny and [lack of] paternal care) are better predictors of speciation than sexual dichromatism. No bioclimatic measures of environmental variability (temperature seasonality, long-term temperature variation and spatial temperature variation) predicted speciation, but range size was negatively associated with two measures of speciation ($\lambda_{DR}$, $\lambda_{ND}$). This suggests that species with smaller range sizes have increased speciation rate, however, small range size may be a cause or effect of elevated speciation rate and potentially even an artifact of taxonomic classification. Interestingly we found a discrepancy in the model results obtained from tip-rate statistics ($\lambda_{DR}$, $\lambda_{ND}$) and those using tip-rate estimates from a model-based approach ($\lambda_{BAMM}$); whereby BAMM produced precisely null results for most associations between predictors and speciation as well as extinction.

The association between sexual selection and sexual dichromatism in birds is a Darwinian trope [@Darwin_1871] that allows it to be commonly used as a proxy for the strength of sexual selection in comparative studies [*e.g.*, @Barraclough_1995; @Seddon_2013; @Morrow_Speciation_2003; @Owens_1999; @Huang_2014]. Despite the association between sexual selection and sexual dichromatism in passerine birds [@Dale_2015] we find that sexual dichromatism does not predict speciation rate, whereas male-biased sexual selection does. Given our results, we suggest that sexual dichromatism may not be a robust proxy for sexual selection, and that variance in the association between dichromatism and the strength of sexual selection may obscure any relationship with speciation rate, despite the high power of our comparative study. There are several reasons why the use of sexual dichromatism as a proxy for sexual selection is problematic. Firstly, ecological pressures could drive sexual dichromatism in the absence of mate choice if sex-specific niches provide opportunity for the evolution of sex-limited traits [@Wallace_1889; @Kottler_1980; @Slatkin_1984; @Shine_1989]. Specifically, natural selection on females may be important in explaining patterns of sexual dichromatism in birds [@Price_2014]. In fairy-wrens (*Malurus cyaneus*), female colouration has evolved in response to environmental pressures, leading to increased sexual dichromatism in regions where females are under stronger selection from predation [@Medina_2017]. Ecological impacts on sexual dichromatism are supported by our path analysis, which reveals that sexual dichromatism is positively affected by temperature seasonality (a measure of environmental variation) albeit relatively weakly. Secondly, colour is but one trait and sexual selection may drive the evolution of sex differences in a wide variety of traits used in mate choice or intrasexual competition [@Miles_2018]. And thirdly, not all plumage colouration honestly reflects mate quality. A recent meta-analysis found that converted carotenoids (but not those sourced from diet) predict mate quality through improvements in parasite resistance and reproductive success [@Weaver_2018]. Given the importance of honest signaling in sexual selection and the variability in colour production across birds, sexual dichromatism in many species may be an unreliable measure of mate quality for female birds and sexual selection for researchers.

In-line with predictions and a previous meta-analysis [@Kraaijeveld_2011] we found that male-biased sexual selection increases speciation rate --- at least for $\lambda_{DR}$ and $\lambda_{ND}$ estimates --- but this association was not dependent on variable or stressful environments. When evolving under stressful conditions, sexually antagonistic selection is hypothesised to be weakened relative to selection that is concordant between the sexes [@Connallon_SA_2015; @Connallon_Hall_2016]. While these predictions are supported in several lab (experimental evolution) studies [@Long_2012; @Berger_2014; @Punzalan_2013], other studies have not found the association [@Delcourt_2009; @Martinossi-Allibert_Savković_2017; @Holman_Jacomb_2017]. The environmental variability predictors used here often reflected changes in climate across space and time, whereby high environmental variation would likely increase the strength of natural selection on a given species. Given that we did not detect an effect of environmental variability (by itself or as an interaction with sexual selection) we are left with several possible explanations. Firstly, the effects of sexual selection on adaptation and thus speciation may depend on the type of environmental variability the species is evolving under. Specifically, sexual antagonism may be more easily purged in environments where cyclic pressures (*e.g.* seasonality) are perpetuated; whereas when environmental variability is directional (*e.g.* long-term climatic change) sexual antagonism can remain indefinitely as a fitness burden [@Connallon_Hall_2016]. Secondly, the environmental predictors used here may not account for the key ecological forces/natural selection pressures that interact with sexual selection to drive speciation. Specifically, access to dietary resources, and the impacts of predation or parasitism are unaccounted for here: arguably these are key processes affecting sexual selection and speciation [*reviewed in* @Maan_2011]. Thirdly, it is also possible that there is no effect of environmental variability on speciation rates; at least not in the species investigated here.

@@

The most consistent finding uncovered in this comparative analysis is that smaller range sizes are associated with increased speciation rates; however reduced range size can be a cause or consequence of speciation. Intuitively, large range size should promote speciation by creating greater opportunities for geographic barriers to form [@Rosenzweig_1995]. However, birds with smaller range sizes are predicted to have lower dispersal ability. This ensues that individuals might not be able to overcome geographical barriers and maintain gene flow; which would promote reproductive isolation and speciation [@Birand_2012; @Claramunt_2012]. Several studies have used morphological measures of wing shape (Hand-wing index/ Kipp's distance) --- a proxy for dispersal ability --- as predictors of speciation and coexistence [*e.g.*, @Kennedy_2016; @Pigot_2016]. But currently, published datasets of these measurements are not available for many passerines. Alternatively, reduced range size may be a consequence of speciation. Under this view, high speciation rate may lead to smaller range sizes as niche filling by recently diverged species will suppress the expansion of newly speciated relatives. Within Himalayan songbirds, range expansions are prevented by the competition for resources by related species [@Price_2014]; this supression on range expansion is then expected to prevent future speciation events [@Rosenzweig_1995; @Weir_2011]. Unlike Himalayan songbirds, species undergoing adaptive radiation in new niches are unlikely to be limited by competition for resources from existing taxa. Across islands we expect to see a correlation between speciation and small range size because they reflect newly formed environments with empty niches. Thus, our findings may be heavily dependent on species whose range size is restricted to a single island within an archipelago (*e.g.* *Geospiza* sp.; a.k.a. Darwin's finches). One further explanation for the negative association between range size and sexual dichromatism/sexual selection is the potential bias of taxonomic classification, whereby over-splitting of species in clades with large ranges leads to increased recent phylogenetic branching as well as smaller ranges. Given the problems of determining causality and mode of speciation the reason for the negative association found here remains obscured.
